TSNE Introduction

### 2021-08-21 PaperDraft.Rmd
adding changes to some docs

Despite those differences, tools like t-SNE and MDS are succeptible to abberant data values, much in the same way that a regression model might be succeptible to outliers and influential points.  In a regression model, the estimated coefficients or predictions might be affected in meaningful fashion by the inclusion or exclusion of a single point - such points are referred to as influential (Cook, 1977).  A suite of tools have been developed to identify and measure the impact of influential points on the results of regression models.

# High-dimensional Data Visualization
## What is t-SNE
## Problems with using t-SNE
## Whitening/Pre-PCA

@@

# Defining “Influence” in High Dimensional Maps
## Influence in Regression - Cook’s Distance
## Influence in

# Methods
## Sparse t-SNE
## Cross Validation and Mantel-esque Test

>>>>

Despite those differences, tools like t-SNE and MDS are susceptible to aberrant data values, much in the same way that a regression model might be susceptible to outliers and influential points.  In a regression model, the estimated coefficients or predictions might be affected in meaningful fashion by the inclusion or exclusion of a single point - such points are referred to as influential (Cook, 1977).  A suite of tools have been developed to identify and measure the impact of influential points on the results of regression models.

The goal of this method is to develop a diagnostic that can be used to help identify influential points and potentially quantify how much they impact the resulting mapping, irrespective of how that mapping was created.  Additionally, this method can be used to assess the sensitivity of each method to influential points and the impact that sensititivy might have on how an end user might want to interpret the resulting map produced.  Note that while robustness to influential points might at first seem like a positive trait for an ordination method, masking distinctly different observations in a mapping may have negative consequences as well.

This document overviews the statistical methodology we have developed to identify influential points that, when excluded from a dataset, can meaningfully impact the shape of an ordination created from data. We overview our methodology step by step.

# High-dimensional Data Visualization with t-SNE

One of the notable new methods for high-dimensional data visualization is t-distributed Stochastic Neighbor Embedding, or t-SNE (Van der Maaten, 2011). Since the time that Laurens Van der Maaten published the first t-SNE paper in 2011, it has been cited more than 21,000 times in use cases ranging from biology (Amir et al, 2013) and genetics () to economics (https://economics.yale.edu/sites/default/files/files/Faculty/Tsyvinski/tSNE_draft15.pdf) to natural language processing (https://aneesha.medium.com/using-tsne-to-plot-a-subset-of-similar-words-from-word2vec-bb8eeaea6229) and machine learning.  The method's prevalence is astounding - it is used not only in academia but in industry problems, including tools like FoodGenius (https://github.com/foodgenius/tsneplot).

The method centers on calculation of a balance between two similarity metrics. The first, called $p_{ij}$,is a conditional probability based on the Euclidean distance between pairs of points in the high-dimensional space. In the lower-dimensional representation, this similarity is defined as $q_{ij}$.

A good representation of the high-dimensional space should mean that $p_{ij}$ and $q_{ij}$ look reasonably similar to each other; since these are both simply joint probabilities, the objective function of t-SNE is simply a symmetricized Kullbeck-Leibler divergence between the two, defined below:

$$C = \sum_{i} \sum_{j} p_{j|i} log(\frac{p_{j|i}}{q_{j|i}})$$

t-SNE is an improvement over Stochastic Neightbor Embedding (SNE), which makes use of Gaussian distributions to model both $p$ and $q$ make preserving local structure somewhat difficult. Most of the minimization of objective functions is done via gradient descent (and some methods require simulated annealing because the algorithm can get caught in local optima). Critically, because the optimization is non-convex, and as a consequence of the gradient descent methods used to optimize it, t-SNE has the potential to provide different maps when run on the same data at the same perplexity.

## Problems with using t-SNE

Although t-SNE is highly effective at creating 2 (or 3) dimensional ordinations of high-dimensional data, the mathematical

Finally, performing t-SNE on large datasets can be computationally slow, necessitating a random-walk version that is discussed in a later paper by van der maaten (2014).

## Reduction of Initial Dimensionality with PCA

Additionally, t-SNE is subject to a pre-processing step involving principal component analysis (PCA).  In most implementations, "whitening" via PCA is performed on the initial dataset (or input distance matrix) to reduce the dimensionality to a reasonable value.  For instance, in Van der Maaten's initial paper, several examples reduced the initial dimensionality of the datato 30 for computational gains (Van der maaten, 2011).

Depending on the size of the initial dimensionality of the data, this step may drastically reduce the number of dimensions that t-SNE needs to deal with.  However, it has some potential to impact the resulting ordination, particularly if the number of initial dimensions kept is relatively low or PCA does not do a great job of fitting the original dataset. At best, the t-SNE mapping is only as good as the initial PCA is.  For the simulations shown in this paper, no whitening was performed, and the original distance matrix was passed to t-SNE directly.

@@

# Defining “Influence” in High Dimensional Maps

## Influence in Regression - Cook’s Distance

In regression models, the concept of influential points is

## Sparse t-SNE
## Cross Validation and Mantel-esque Test

## Influence in

# Methods

This document assesses the effect of multivariate outliers on the maps produced by two methods: classical multidimensional scaling and t-SNE. In general, we seek to identify either **multivariate outliers** (or groups of them) or alternatively single influential points that, for some reason, cause large differences in the shape of the resulting ordination.

Our Permanova-based approach is as follows:

1. For each observation in the dataset, remove it, and generate a t-SNE map in 2 dimensions from the resulting (n-1) observations.  (For accounting purposes, it may make sense to keep the nx1 structure but replace the holdout observation with NA). 
2. Generate a distance matrix of the nx1 by 2 matrix. Do this for each of the held-out observations. 
3. Generate a pairwise-complete correlation matrix from the combination of these vectorized distances that measures similarity of distances (where available).
4. Convert that correlation matrix (of all the hold-out vectorized distance matrices) into a distance matrix using a $1-|\sqrt{}\. 
5. Apply non-parametric PERMANOVA tests for each holdout observation - this means that on a 100-observation dataset, you apply 100 individual tests where each test uses an independent variable that is equal to 1 for the index of the holdout observation and 0 elsewhere. The hypotheses are specified as such:

+ $H_0$: No differences in map with observation j removed vs. those without j removed 
+ $H_a$: Some difference in map with observation j removed

We will discuss correcting p-values for multiple testing later, although currently our plan is to assess using Benjamini-Hochberg False Discovery Rate or Bonferroni corrections (or something similar). We would expect that if the maps are suitably distorted, or *influenced* by a single point, the resulting p-value for the jth adonis test would be small.

# Results

# Discussion

t-SNE is not particularly sensitive to single points that are highly dissimilar to others in the dataset. It is often not obvious that such observations are present when viewing a t-SNE map. In the case of other methods, like MDS, influential points change the configuration of the ordination significantly.

### 2021-08-21 tSNE Introduction Document.Rmd
adding changes to some docs

>>>>

One of the notable new methods for high-dimensional data visualization is t-distributed Stochastic Neighbor Embedding, or t-SNE (Van der Maaten, 2011). Since the time that Laurens Van der Maaten published the first t-SNE paper in 2011, it has been cited more than 21,000 times in use cases ranging from biology (Amir et al, 2013) and genetics () to economics (https://economics.yale.edu/sites/default/files/files/Faculty/Tsyvinski/tSNE_draft15.pdf) to natural language processing (https://aneesha.medium.com/using-tsne-to-plot-a-subset-of-similar-words-from-word2vec-bb8eeaea6229) and machine learning.  The method's prevalence is astounding - it is used not only in academia but in industry problems, including tools like FoodGenius (https://github.com/foodgenius/tsneplot).

The method centers on calculation of a balance between two similarity metrics. The first, called $p_{ij}$,is a conditional probability based on the Euclidean distance between pairs of points in the high-dimensional space. In the lower-dimensional representation, this similarity is defined as $q_{ij}$.

A good representation of the high-dimensional space should mean that $p_{ij}$ and $q_{ij}$ look reasonably similar to each other; since these are both simply joint probabilities, the objective function of t-SNE is simply a symmetricized Kullbeck-Leibler divergence between the two, defined below:

$$C = \sum_{i} \sum_{j} p_{j|i} log(\frac{p_{j|i}}{q_{j|i}})$$

t-SNE is an improvement over Stochastic Neightbor Embedding (SNE), which makes use of Gaussian distributions to model both $p$ and $q$ make preserving local structure somewhat difficult. Most of the minimization of objective functions is done via gradient descent (and some methods require simulated annealing because the algorithm can get caught in local optima). Critically, because the optimization is non-convex, and as a consequence of the gradient descent methods used to optimize it, t-SNE has the potential to provide different maps when run on the same data at the same perplexity.
